Generative Dynamics of Supreme Court Citations: Analysis with a New Statistical Network Model

### 2020-08-20 Tex/PA_rr/main.tex
expanded on covariate term write up

We first discuss the dependence effect estimates in Figure \ref{SCC_results_1}. In most of the years that it is estimable, the reciprocity effect is statistically significantly positive, and substantial in magnitude. In a typical year, the presence of a citation from case $i$ to case $j$ increases the log odds of a citation from case $j$ to case $i$ by approximately 0.50. Both transitivity effects---different term transitivity and gwesp---are estimated to be positive and statistically significant in every year, and are even greater in magnitude than the reciprocity effect. The log odds of a citation from case $i$ to case $j$ increases by more than 2 if there is at least one third case, $k$ that is cited by $i$, and cites to $j$. Results regarding the popularity effect (i.e., gwidegree) are generally supportive of the hypothesis of a popularity effect in Supreme Court citations, as the gwidegree estimate is negative and statistically significant at conventional levels in nearly every term. This effect looks to have grown substantially after the 2000 term, which also coincides closely with the beginning of the Roberts court in 2006.

@@

The effects of the exogenous covariates included in the c-ERGM are presented in Figure \ref{SCC_results_2}.  Most of the exogenous covariate effects align with expectations. We find that cases are more likely to cite those that (1) have been decided most recently, (2) were decided by large majorities, (3) have not been overruled, (4) were authored by the same justice, and (5) are in the same issue area. Surprisingly, we do not find consistent effects of either of the covariates that involve justice ideology. The ideological distance between two cases, as measured by the absolute difference in the median of the majority coalitions; and the ideological breadth of the cited cases, as measured by the absolute difference between the most liberal and most conservative justices on the potential receiving case, both exhibit signs and significance levels that are highly volatile. Our results leave open the question of whether, and if so, how, citations are shaped by ideological factors.

>>>>

\subsubsection{Dependence Terms}
We first discuss the dependence effect estimates in Figure \ref{SCC_results_1}. In most of the years that it is estimable, the reciprocity effect is statistically significantly positive, and substantial in magnitude. In a typical year, the presence of a citation from case $i$ to case $j$ increases the log odds of a citation from case $j$ to case $i$ by approximately 0.50. Both transitivity effects---different term transitivity and gwesp---are estimated to be positive and statistically significant in every year, and are even greater in magnitude than the reciprocity effect. The log odds of a citation from case $i$ to case $j$ increases by more than 2 if there is at least one third case, $k$ that is cited by $i$, and cites to $j$. Results regarding the popularity effect (i.e., gwidegree) are generally supportive of the hypothesis of a popularity effect in Supreme Court citations, as the gwidegree estimate is negative and statistically significant at conventional levels in nearly every term. This effect looks to have grown substantially after the 2000 term, which also coincides closely with the beginning of the Roberts court in 2006.

@@

\subsubsection{Covariate Terms}
Effects of the exogenous covariates included in the c-ERGM are presented in Figure \ref{SCC_results_2}. Most of the exogenous covariate effects align with expectations and have relatively stable parameter estimates over time. We find that cases are more likely to cite those that (1) have been decided most recently, (2) were authored by the same justice, (3) are in the same issue area, and (4) have not been overruled. Surprisingly, we do not find consistent effects for majority coalition size or for the two covariates that involve justice ideology. The effect of majority coalition size was generally not statistically significant until the later portion of the period examined; ideological distance between two cases had signs and significance levels that are highly inconsistent; and the ideological breadth of the potentially cited case was effectively zero for the entire period.

These results leave open the question of whether and how citations are shaped by ideological factors. While we did not formally hypothesize about temporal dynamics in the effects, a closer look at the observed trends allows us to identify some patterns that suggest directions for future research in this area. In this discussion, we focus on the shift in the effect of ideological distance beginning with the Rehnquist court and extending into the Roberts court.

The existing body of evidence indicates that chief justices impact the courts they preside over beyond casting one of the nine votes \citep{cross2005decisional,danelski2016chief}. It is therefore within expectations that there are discernible differences in citation patterns across different courts. With regard to the effect of ideological distance, we see that the Vinson (1950-1953) and Warren (1954-1969) courts were highly volatile, the Burger court (1970-1986) had a generally higher likelihood of citing opinions that were ideologically closer to itself, and the more recent Rehnquist (1987-2005) and Roberts (2006-2015) courts showed some tendency toward citing cases that are ideologically distant. The latter finding might be initially puzzling, but can be explained in light of observations from legal scholars that the Rehnquist court was characterized by a `split-the-difference' jurisprudence toward its later half \citep{wilkinson2005rehnquist,basiak2006roberts}. This means that the court strove to take moderate positions that compromised on both sides of the debated issues, which in large differed from the jurisprudential approach of prior courts. Less has been written about the Roberts court on this topic, but Roberts shares many similarities with Rehnquist especially in comparison to previous chief justices, and has himself noted the desirability of widely accepted rulings even at the expense of their scope \citep{pomerance2018center,sunstein2008trimming}.

One way to achieve the kind of balance required for support is to cite broadly from both sides of the argument \citep{wilkinson2005rehnquist}, which increases the likelihood of citing ideologically distant cases. Consider for example Hamdi v. Rumsfeld (542 U.S. 1, 2004) which ruled on the question of whether a U.S. citizens can be detained as an ``enemy combatant''. The court, in a four-justice plurality opinion supported by two additional concurring justices, decided at the same time that ``a state of war is not a blank check for the President'' and the due process assessment must ``pay keen attention to the particular burdens faced by the Executive in the context of military action''. In total, the plurality opinion in Hamdi v. Rumsfeld cited 47 opinions, with cites as distant as 2.42 and 2.13 on either side of its median position.

In the above discussion, we identified an explanation for the observed temporal dynamics in the effect of ideological distance. Specifically, ideologically distant cases become more important as courts adopt the relatively new `split-the-difference' jurisprudential approach which requires broad citations. We recognize that there are alternative explanations---perhaps one that is related to the coterminous change in the effect of majority coalition size---but a deeper look into the relevant literature is beyond the scope of this paper. Our discussion serves to illustrate that term-by-term c-ERGMs can uncover important temporal dynamics that courts researchers can use to study a much greater range of citation-related phenomenon beyond what we present here. To the extent that the US Supreme Court is a political entity that interacts with the broader society, the body of court opinions and citations between them is corpus that continues to grow based on both its history and its contemporary context. The methods we present here afford researchers the tools to better understand these complexities.
